Meta-analytic approaches and effect sizes to account for 'nuisance heterogeneity' in comparative physiology

### 2021-07-25 MS/ms.Rmd
Read and fixed a few small things in MS

Meta-analysis is a powerful tool to generate quantitatively informed answers to pressing global challenges. By distilling data from broad sets of research designs and study systems into standardized effect sizes, meta-analyses provide physiologists with opportunities to estimate overall effect sizes and understand the drivers of effect variability. Despite this ambition, research designs in the field of comparative physiology can appear, at the outset, as being vastly different to each other because of 'nuisance heterogeneity' (e.g., different temperatures or treatment dosages used across studies). Methodological differences across studies have led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with multilevel meta-regression models to both account for the factors driving differences across studies and make them more comparable. We formalise effect size measures (e.g., $Q_{10}$) that provide comparative physiologists with a means to remove nuisance heterogeneity without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be applied to a variety of different contexts to derive new effects sizes and sampling variances opening up new possibilities. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions and make results from large-scale data sets more accessible, comparable and widely interpretable.

@@

Meta-analysis has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is increasingly prominent in the field of comparative physiology  [@Gurevitch2018; @Nakagawa2017; @Glass2015]. Meta-analyses answer research questions by synthesising research results and identifying sources of variation across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Arnqvist1995; @Koricheva2013; @Borenstein2019]. Ideally, meta-analyses are part of a systematic review and methods are carefully reported upon to ensure the review and data collation are transparent and reproducible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolutionary Biology (PRISMA-EcoEvo) checklist, @ODea2021].

A meta-analysis can have three goals: 1) to provide an overall mean estimate of a treatment effect or relationship, 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies and 3) attempt to identify research gaps and publication biases [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013; @Borenstein2019]. Meta-analysing independent studies provides greater statistical power and precision than what any individual study on its own would be able to provide - particularly given that most empirical studies are already under-powered in many areas of biology [@Forstmeier2017; @Jennions2003; @Button2013]. By expressing study effects on a common scale (i.e., standardised effect size) we can gain broader insight into the direction and efficacy of a particular treatment effect or the strength of a relationship between two variables of interest [@Gurevitch2018; @Koricheva2013]. Meta-analyses have already provided comparative physiologists with powerful insights on pressing global challenges - from testing whether physiological plasticity can buffer organisms against climate change [e.g, @Seebacher2015] to the degree to which endocrine disrupting chemicals, such as bis-phenol A (BPA), impact aquatic organisms [e.g., @WuSeebacher2020].

@@

The goal of our review is to briefly overview the meta-analytic process and argue for the value of heterogeneity in answering fundamental research problems and guiding research directions. We show how 'nuisance heterogeneity' in comparative physiology can be comfortably dealt with by re-formalising many existing effect size measures and / or by using multilevel meta-regression models. Then, we present results from a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by nuisance heterogeneity and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. We also formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists with opportunities to explicitly incorporate 'nuisance heterogeneity' at the effect size level to ease their interpretation. Finally, we describe how more complex treatment differences, such as non-linear dosage differences, can be accommodated using multilevel meta-regression models. We hope that, by expanding the meta-analytic toolkit, it will provide new opportunities for comparative physiologists to address how organisms will cope with rapidly changing environments and anthropogenic stressors in the future.

@@

We tend to agree with @Glass2015, and his sentiment is particularly relevant to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms, including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are different. Synthesizing diverse study outcomes allows us to also understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction), and approaches (e.g., short-term vs long-term studies) that can explain the diversity of effect outcomes observed. Doing so provides a rich set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included in a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility (or ideally should be) are such a prominent feature of meta-analysis [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogeneity then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogeneity, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant and may even be misleading. It may, therefore be more useful to understand overall effects in each of these groups separately, and assess the extent to which they differ.

@@

The results of our literature survey show that quantitative syntheses in comparative physiology commonly analyse raw summary statistics (`r with(effects, effects[effect == "Means","prop"])*100`% of studies), followed closely by Fishers transformed correlation coefficients (*Zr*) (`r with(effects, effects[effect == "Zr","prop"])*100`% of studies), standardised mean differences (*SMD*) (`r with(effects, effects[effect == "SMD","prop"])*100`%  of studies), and log response ratios (*lnRR*) (`r with(effects, effects[effect == "lnRR","prop"])*100`%, of studies) (Figure \@ref(fig:figsurv)A). In many cases, studies make use of multiple effect size measures. Approximately `r with(hetero, hetero[cont_moderator_variable == "yes","percent"])*100`% of effect sizes used within the study we deemed to be impacted by 'nuisance heterogeneity' in some form or another (Figure \@ref(fig:figsurv)B), but of the studies where this was the case, only `r with(hetero_dealt, hetero_dealt[cont_moderator_variable_modelled == "yes","percent"])*100`% of the studies clearly dealt with it (Figure \@ref(fig:figsurv)B), often through meta-regression modelling. Overall, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`% of studies used a weighted meta-analytic model; however, of the studies self described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not account for sampling variance. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:figsurv)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity, which is unfortunate given how important these measures are to interpreting effects.

@@

```{r tablea, echo = FALSE, tab.cap = "Common effect sizes used throughout meta-analyses in comparative physiology, their associated sampling variances and examples on when they might be used. M: mean; SD: standard deviation; N: sample size; CTmin: critical thermal minimum; CTmax: critical thermal maximum; LC_50: median lethal concentration; LT_50: median lethal temperature; BPA: bisphenol A."}

@@

# Note: It is esssential that you be clear that the function 'compose' is exported from the flextable package. That's because purr also has the same function, so it gets messy.
tablea <- flextable(table_fin)                                                                                %>% 
          flextable::compose(j = 1:4, part = "header", value = as_paragraph(as_b(c("Effect Measure", "Definition", "Sampling Variance", "Examples")))) %>% 
          flextable::compose(j = 2:3, value = as_paragraph(as_equation(c( eqns_yi, eqns_vi))))             %>%
          flextable::compose(j = 4, value = as_paragraph(eg))             %>%
          flextable::width(j = 1, 3)                                                                          %>% 
          flextable::width(j = 2, 2) %>% 
          flextable::width(j = 4, 5) %>% 
          flextable::footnote(i = 4, j = 1, value = as_paragraph(as_equation(footnotes)),
               ref_symbols = c("a"),
               part = "body", inline = TRUE)

@@

As can be seen in Table \@ref(tab:tablea), none of the effect measures control for 'nuisance heterogeneity'. For example, the magnitude of *lnRR* or *SMD* might depend on differences in the temperatures, dosages, pH, or water potential applied to each treatment. The same applies to modelling the means. For example, standard metabolic rate (SMR) is temperature-dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017] (if, of course, temperature is not of interest itself). While we could model this explicitly, this may not always be desirable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling. In Box 1, we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in Box 1 that can be easily generalized to alternative effect size measures and cases. Below, we show how we can formalise a traditional effect size, such as $lnRR$, to capture this ‘nuisance variation’ using a common, well understood effect measure of interest to comparative physiologists ($Q_{10}$). Figure \@ref(fig:fig0) summarises different effect sizes that we mention in Table 1, Box 1 and the section below.

```{r, fig0, fig.cap="**Effect sizes from old to new**. ‘Traditional’ effect sizes (top left) are somehow detached from more biologically relevant measures commonly used in ‘Ccomparative physiology’ (top right) although statistical connections have always existed (‘Universal’; top middle). By combining traditional effects with physiological parameters, and correcting the resulting quantities for nuisance parameters, we can derive a new generation of biologically relevant, easily interpretable effect measures for use in meta-analyses (‘Expanded’; bottom). For abbreviations, see Table 1."}
image <- image_read(here::here("figs", "Fig0_v1_ML.png"))
image
```

@@

In equation \@ref(eq:lnRR), $M_{1}$ is the mean of group 1 (e.g., a control group), where as $M_{2}$ is the mean of group 2 (e.g., a treatment group). The mean, $M_{i}$, for group *i* can be any measurement type (e.g. a physiological rate, mass etc) so long as the variable is measured on the ratio scale. Natural log transformation of this ratio makes the effect size normally distributed (a commonly assumed by meta-analytic models). Equation \@ref(eq:slnRR) is the analytical solution for *lnRR*'s sampling variance where $SD_{1}^2$ and $SD_{2}^2$ are the sample variance in group 1 and 2, respectively, and $N_{1}$ and $N_{2}$ are the sample size in group 1 and 2.

@@

It will not be possible to always derive an effect size that completely controls for nuisance heterogeneity at the effect size level all the time. This limitation is partly because treatment effects may not be simple linear or exponential functions of the mean response, making the effect sizes we discuss above unsuitable [e.g., thermal performance curves; @Noble2018]. For example, dosages applied to treatments can follow quite complex non-linear relationships in relation to some mean response (e.g., quadratic relationships). When this occurs, we recommend comparative physiologists to apply meta-regression approaches, controlling for nuisance heterogeneity as a moderator. In many cases, this will be the easiest, or even preferred, approach because exploring drivers of heterogeneity is of prime interest in most meta-analyses anyway. Nonetheless, there are some simple statistical approaches a comparative physiologist can use to make the overall effects more easily interpretable, allowing one to control for nuisance heterogeneity, while also testing other biological moderators of interest [@Schielzeth2010]. Before diving into meta-regression models, we will first overview the meta-analytic models commonly used.

@@

where $\sum \beta_{j}x_{j}$ is simply the sum of all effects for all moderators (fixed effects), $x_{j}$. All other notation is the same as described above. Importantly, the variable $x_{j}$ could be any moderator collected from studies, including dosage, temperature, salinity, or pH. Some moderators will be at the level of the study (e.g., was the study experimental or observational in nature) but others could be taken at the effect size level (i.e., was the effect derived using a temperature of 23°C or 35°C). By including these moderators in a meta-regression model, the interpretation of the overall meta-analytic mean, $\beta_{0}$, changes. For example, assume we collected the change in *in ovo* testosterone dosage relative to a control group (i.e., in ng T / g yolk) for our meta-analysis on egg T effects on bird offspring traits in birds. If the untransformed dosage, $x_{j}$, is included in the model then the overall meta-analytic mean estimate, $\beta_{0}$, would be the mean effect of T on reproduction when the dosage difference between the treatment and control group is 0 ng T / g yolk (i.e., it would be reflective of unmanipulated eggs). Of course, this interpretation does not make much sense.

We can, however, apply some simple transformations to make the overall mean estimate more intuitive while also improving the interpretation of non-linear parameters and interactions estimated from the model [@Schielzeth2010; @GelmanHill2007]. One simple transformation would be to centre the variable $x_{j}$ by subtracting each value from the mean to create a new input variable (i.e., $c_{j}=x_{j}-\bar{x}$, where the subscript *j* denotes each individual effect size dosage). The new variable, $c_{j}$, is now centered on the mean and can replace $x_{j}$ in the meta-regression model. When the model is refit with $c_{j}$, the estimated $\beta_{0}$ can now be more intuitively interpreted as the overall mean effect of CORT on reproduction at an average dosage difference between treatment and control groups. Importantly, by centering, we retain the original units (ng T / g yolk) of the variable, which can be very useful to ease the interpretation of the effect magnitude (e.g., when the dosage difference is 10 ng T / g yolk). Given it is very common to estimate and compare overall meta-analytic means (i.e., $\beta_{0}$) in meta-analyses, this centering can be particularly useful in making these means more interpretable. Mean centering can even be extended to include other 'nuisance' variables that might creep into the dataset; the same centering can be done with a second variable, and the overall mean adjusted and interpreted in the context of this new variable along with dosage. Of course, if one is not interested in the magnitude of the effect, we can simply model mean reproductive output as a function of dosage (often what are referred to as 'arm-based' meta-analytic models). Here, one can model the mean for each group (i.e., control and treatment) and account for the known sampling variance associated with each group mean. Dosage, or other continuous variables, like temperature, can then be modeled (accounting for sampling variance) to understand how mean reproductive output varies. It is important to recognise that all means need to be comparable and that these models are more complex to fit. While they can be equivalent to 'contrast-based' models, they require complex interactions to be estimated [@Nakagawa2015].

@@

As we have already emphasised, quantifying effect size heterogeneity provides the impetus for exploring why effects might vary within and across studies, while also providing context for interpreting the generality of effects. There are a number of measures commonly used to quantify effect size heterogeneity [@Borenstein2019]; however, in ecological and evolutionary meta-analysis, heterogeneity is often calculated and reported as $I_{total}^2$ as follows (if using a multilevel meta-analytic model as we describe above):

@@

Comparative physiologists are interested in meta-analysing effects across a diversity of species, experimental designs, and environments. Importantly, sampling variances for many effect sizes commonly used by comparative physiologists (e.g., CTmax, $Q_{10}$ etc) can be formalized and powerful meta-analytic models can be easily applied to these effect measures (Figure \@ref(fig:fig0)). While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogeneity will be the primary interest. However, factors such as temperature and dosage differences across studies may generate obvious sources of heterogeneity that may not be of prime interest.  Instead, comparative physiologists might simply want to focus on biological drivers of effect size variability and want an effect size or analytic approach that allows them to remove these ‘nuisances’. Here, we provide a set of tools to deal with ‘nuisance heterogeneity’ (Figure \@ref(fig:fig1))by reinventing standard effect sizes and/or using mean centered nuisance variables in meta-regression models. Estimating complex meta-regression models with a multitude of fixed effects might result, however, in more challenging model interpretation. As such, a combined approach, where some of the known heterogeneity is dealt with by using an appropriate effect size and associated sampling variance while other biologically interesting moderators are fit in a meta-regression model might be desired. Indeed, this may even be a necessity to simplify the model. The set of tools we describe provide clarification around importance and limitations of heterogeneity. We also hope to provide new ways to help comparative physiologists communicate effect measures more richly to guide our understanding and decisions around pressing global challenges.

>>>>

Meta-analysis is a powerful tool to generate quantitatively informed answers to pressing global challenges. By distilling data from broad sets of research designs and study systems into standardized effect sizes, meta-analyses provide physiologists with opportunities to estimate overall effect sizes and understand the drivers of effect variability. Despite this ambition, research designs in the field of comparative physiology can appear, at the outset, as being vastly different to each other because of 'nuisance heterogeneity' (e.g., different temperatures or treatment dosages used across studies). Methodological differences across studies have led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with multilevel meta-regression models to both account for the factors driving differences across studies and make them more comparable. We formalise effect size measures (e.g., $Q_{10}$) that provide comparative physiologists with a means to remove nuisance heterogeneity without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be applied to a variety of different contexts to derive new effects sizes and sampling variances opening new possibilities. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions and make results from large-scale data sets more accessible, comparable and widely interpretable.

@@

Meta-analysis has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is becoming increasingly prominent in the field of comparative physiology  [@Gurevitch2018; @Nakagawa2017; @Glass2015]. Meta-analyses answer research questions by synthesising research results and identifying sources of variation across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Arnqvist1995; @Koricheva2013; @Borenstein2019]. Ideally, meta-analyses are part of a systematic review and methods are carefully reported upon to ensure the review and data collation are transparent and reproducible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolutionary Biology (PRISMA-EcoEvo) checklist, @ODea2021].

A meta-analysis can have three goals: 1) to provide an overall mean estimate of a treatment effect or relationship, 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies and 3) attempt to identify research gaps and publication biases [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013; @Borenstein2019]. Meta-analysing independent studies provides greater statistical power and precision than what any individual study on its own would be able to provide -- particularly given that most empirical studies are already under-powered in many areas of biology [@Forstmeier2017; @Jennions2003; @Button2013]. By expressing study effects on a common scale (i.e., standardised effect size) we can gain broader insight into the direction and efficacy of a particular treatment effect or the strength of a relationship between two variables of interest [@Gurevitch2018; @Koricheva2013]. Meta-analyses have already provided comparative physiologists with powerful insights on pressing global challenges -- from testing whether physiological plasticity can buffer organisms against climate change [e.g, @Seebacher2015] to the degree to which endocrine disrupting chemicals, such as bis-phenol A (BPA), impact aquatic organisms [e.g., @WuSeebacher2020].

@@

The goal of our review is to briefly overview the meta-analytic process and argue for the value of heterogeneity in answering fundamental research problems and guiding research directions. We present results from a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by nuisance heterogeneity and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. Then, we show how 'nuisance heterogeneity' in comparative physiology can be comfortably dealt with by re-formalising many existing effect size measures and / or by using multilevel meta-regression models. We formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists with opportunities to explicitly incorporate 'nuisance heterogeneity' at the effect size level to ease their interpretation. Finally, we describe how more complex treatment differences, such as non-linear dosage differences, can be accommodated using multilevel meta-regression models. We hope that, by expanding the meta-analytic toolkit, it will provide new opportunities for comparative physiologists to address how organisms will cope with rapidly changing environments and anthropogenic stressors in the future.

@@

We tend to agree with @Glass2015, and his sentiment is particularly relevant to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms, including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are different. Synthesizing diverse study outcomes allows us to also understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction), and approaches (e.g., short-term vs long-term studies) that can explain the diversity of effect outcomes observed. Doing so provides a rich set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included in a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogeneity then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogeneity, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant and may even be misleading. It may, therefore be more useful to understand overall effects in each of these groups separately, and assess the extent to which they differ.

@@

The results of our literature survey show that quantitative syntheses in comparative physiology commonly analyse raw summary statistics (`r with(effects, effects[effect == "Means","prop"])*100`% of studies), followed closely by Fishers transformed correlation coefficients (*Zr*) (`r with(effects, effects[effect == "Zr","prop"])*100`% of studies), standardised mean differences (*SMD*) (`r with(effects, effects[effect == "SMD","prop"])*100`%  of studies), and log response ratios (*lnRR*) (`r with(effects, effects[effect == "lnRR","prop"])*100`%, of studies) (Figure \@ref(fig:figsurv)A). In many cases, studies make use of multiple effect size measures. Approximately `r with(hetero, hetero[cont_moderator_variable == "yes","percent"])*100`% of effect sizes used within the study were deemed to be impacted by nuisance heterogeneity in some form or another (Figure \@ref(fig:figsurv)B), but of the studies where this was the case, only `r with(hetero_dealt, hetero_dealt[cont_moderator_variable_modelled == "yes","percent"])*100`% of the studies clearly dealt with it (Figure \@ref(fig:figsurv)B), often through meta-regression modelling. Overall, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`% of studies used a weighted meta-analytic model; however, of the studies self described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not account for sampling variance. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:figsurv)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity, which is unfortunate given how important these measures are to interpreting effects.

@@

```{r tablea, echo = FALSE, tab.cap = "**Common effect sizes used throughout meta-analyses in comparative physiology, their associated sampling variances and examples on when they might be used**. M: mean; SD: standard deviation; N: sample size; CTmin: critical thermal minimum; CTmax: critical thermal maximum; LC_50: median lethal concentration; LT_50: median lethal temperature; BPA: bisphenol A."}

@@

# Note: It is esssential that you be clear that the function 'compose' is exported from the flextable package. That's because purr also has the same function, so it gets messy.
tablea <- flextable(table_fin)                                                                                %>% 
          flextable::compose(j = 1:4, part = "header", value = as_paragraph(as_b(c("Effect Measure", "Definition", "Sampling Variance", "Examples")))) %>% 
          flextable::compose(j = 2:3, value = as_paragraph(as_equation(c( eqns_yi, eqns_vi))))             %>%
          flextable::compose(j = 4, value = as_paragraph(eg))             %>%
          flextable::width(j = 1, 3)                                                                          %>% 
          flextable::width(j = 2, 2) %>% 
          flextable::width(j = 4, 5) %>% 
          flextable::footnote(i = 4, j = 1, value = as_paragraph(as_equation(footnotes)),
               ref_symbols = c("a"),
               part = "body", inline = TRUE) %>% 
          flextable::font(part = "all", fontname = "Times New Roman")

@@

As can be seen in Table \@ref(tab:tablea), none of the effect measures control for nuisance heterogeneity. For example, the magnitude of *lnRR* or *SMD* might depend on differences in the temperatures, dosages, pH, or water potential applied to each treatment. The same applies to modelling the means. For example, standard metabolic rate (SMR) is temperature-dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017] (if, of course, temperature is not of interest itself). While we could model this explicitly, this may not always be desirable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling. In Box 1, we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in Box 1 that can be easily generalized to alternative effect size measures and cases. Below, we show how we can formalise a traditional effect size, such as $lnRR$, to capture this nuisance variation using a common, well understood effect measure of interest to comparative physiologists ($Q_{10}$). Figure \@ref(fig:fig0) summarises the different effect sizes that we mention in Table 1, Box 1 and the section below.

```{r, fig0, fig.cap="**Effect sizes from old to new**. ‘Traditional’ effect sizes (top left) are somehow detached from more biologically relevant measures commonly used in ‘Comparative physiology’ (top right) although statistical connections have always existed (‘Universal’; top middle). By combining traditional effects with physiological parameters, and correcting the resulting quantities for nuisance variation, we can derive a new generation of biologically relevant, easily interpretable effect measures for use in meta-analyses (‘Expanded’; bottom). For abbreviations, see Table 1."}
image <- image_read(here::here("figs", "Fig0_v1_ML.png"))
image
```

@@

In equation \@ref(eq:lnRR), $M_{1}$ is the mean of group 1 (e.g., a control group), whereas $M_{2}$ is the mean of group 2 (e.g., a treatment group). The mean, $M_{i}$, for group *i* can be any measurement type (e.g. a physiological rate, mass etc) so long as the variable is measured on the ratio scale. Natural log transformation of this ratio makes the effect size normally distributed (as commonly assumed by meta-analytic models). Equation \@ref(eq:slnRR) is the analytical solution for *lnRR*'s sampling variance where $SD_{1}^2$ and $SD_{2}^2$ are the sample variance in group 1 and 2, respectively, and $N_{1}$ and $N_{2}$ are the sample size in group 1 and 2.

@@

It will not always be possible to derive an effect size that completely controls for nuisance heterogeneity at the effect size level. This limitation is partly because treatment effects may not be simple linear or exponential functions of the mean response, making the effect sizes we discuss above unsuitable [e.g., thermal performance curves; @Noble2018]. For example, dosages applied to treatments can follow quite complex non-linear relationships in relation to some mean response (e.g., quadratic relationships). When this occurs, we recommend comparative physiologists to apply meta-regression approaches, controlling for nuisance heterogeneity as a moderator. In many cases, this will be the easiest, or even preferred, approach because exploring drivers of heterogeneity is of prime interest in most meta-analyses anyway. Nonetheless, there are some simple statistical approaches a comparative physiologist can use to make the overall effects more easily interpretable, allowing one to control for nuisance heterogeneity, while also testing other biological moderators of interest [@Schielzeth2010]. Before diving into meta-regression models, we will first overview the meta-analytic models commonly used.

@@

where $\sum \beta_{j}x_{j}$ is simply the sum of all effects for all moderators (fixed effects), $x_{j}$. All other notation is the same as described above. Importantly, the variable $x_{j}$ could be any moderator collected from studies, including dosage, temperature, salinity, or pH. Some moderators will be at the level of the study (e.g., was the study experimental or observational in nature) but others could be taken at the effect size level (i.e., was the effect derived using a temperature of 23°C or 35°C). By including these moderators in a meta-regression model, the interpretation of the overall meta-analytic mean, $\beta_{0}$, changes. For example, assume we collected the change in *in ovo* testosterone dosage relative to a control group (i.e., in ng T / g yolk) for our meta-analysis on egg T effects on bird offspring traits. If the untransformed dosage, $x_{j}$, is included in the model then the overall meta-analytic mean estimate, $\beta_{0}$, would be the mean effect of T on traits when the dosage difference between the treatment and control group is 0 ng T / g yolk (i.e., it would be reflective of unmanipulated eggs). Of course, this interpretation does not make much sense.

We can, however, apply some simple transformations to make the overall mean estimate more intuitive while also improving the interpretation of non-linear parameters and interactions estimated from the model [@Schielzeth2010; @GelmanHill2007]. One simple transformation would be to centre the variable $x_{j}$ by subtracting each value from the mean to create a new input variable (i.e., $c_{j}=x_{j}-\bar{x}$, where the subscript *j* denotes each individual effect size dosage difference). The new variable, $c_{j}$, is now centered on the mean and can replace $x_{j}$ in the meta-regression model. When the model is refit with $c_{j}$, the estimated $\beta_{0}$ can now be more intuitively interpreted as the overall mean effect of T on offspring traits at an average dosage difference between treatment and control groups. Importantly, by centering, we retain the original units (ng T / g yolk) of the variable, which can be very useful to ease the interpretation of the effect magnitude (e.g., when the dosage difference is 10 ng T / g yolk).

Given it is very common to estimate and compare overall meta-analytic means (i.e., $\beta_{0}$) in meta-analyses, this centering can be particularly useful in making these means more interpretable. Mean centering can even be extended to include other 'nuisance' variables that might creep into the dataset; the same centering can be done with a second variable, and the overall mean adjusted and interpreted in the context of this new variable along with dosage. Of course, if one is not interested in the magnitude of the effect, we can simply model mean reproductive output as a function of dosage (often what are referred to as 'arm-based' meta-analytic models). Here, one can model the mean for each group (i.e., control and treatment) and account for the known sampling variance associated with each group mean. Dosage, or other continuous variables, like temperature, can then be modeled (accounting for sampling variance) to understand how mean reproductive output varies. It is important to recognise that all means need to be comparable and that these models are more complex to fit. While they can be equivalent to 'contrast-based' models, they require complex interactions to be estimated [@Nakagawa2015].

@@

As we have already emphasised, quantifying effect size heterogeneity provides the impetus for exploring why effects might vary within and across studies, while also providing context for interpreting the generality of overall effects. There are a number of measures commonly used to quantify effect size heterogeneity [@Borenstein2019]; however, in ecological and evolutionary meta-analysis, heterogeneity is often calculated and reported as $I_{total}^2$ as follows (if using a multilevel meta-analytic model as we describe above):

@@

Comparative physiologists are interested in meta-analysing effects across a diversity of species, experimental designs, and environments. Importantly, sampling variances for many effect sizes commonly used by comparative physiologists (e.g., CTmax, $Q_{10}$ etc) can be formalized and powerful meta-analytic models can be easily applied to these effect measures (Figure \@ref(fig:fig0)). While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogeneity will be the primary interest. However, factors such as temperature and dosage differences across studies may generate obvious sources of heterogeneity that may not be of prime interest.  Instead, comparative physiologists might simply want to focus on biological drivers of effect size variability and want an effect size or analytic approach that allows them to remove these ‘nuisances’. Here, we provide a set of tools to deal with ‘nuisance heterogeneity’ (Figure \@ref(fig:fig1)) by reinventing standard effect sizes and/or using mean centered nuisance variables in meta-regression models. Estimating complex meta-regression models with a multitude of fixed effects might result, however, in more challenging model interpretation. As such, a combined approach, where some of the known heterogeneity is dealt with by using an appropriate effect size and associated sampling variance while other biologically interesting moderators are fit in a meta-regression model might be desired. Indeed, this may even be a necessity to simplify the model. The set of tools we describe provide clarification around importance and limitations of heterogeneity. We also hope to provide new ways to help comparative physiologists communicate effect measures more richly to guide our understanding and decisions around pressing global challenges.
